---
title: "Final Project"
output: html_document
date: "2023-11-30"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}

library(tidyverse)
merged_data <- read_delim("merged_data.csv")
 
```

# **Introduction - Angelic**

SNAP stands for the Supplemental Nutrition Assistance Program in the United States and is implemented and maintained by the government. It has replaced the food stamp system in the U.S. which provides food aid for families and individuals who meet a specific set of income and resource requirements. As food costs rise and wages stagnate, more and more people have been forced to utilize emergency food services. Right now, millions of people across the United States rely on SNAP due to food insecurity. 

Populations that turn to emergency food programs are often the most vulnerable, with little to no income nor the resources to buy nutritious food. Food banks are often stocked with canned, non-perishable items or leftovers from grocery stores like chips, sweets, or ice cream in some cases. Even being eligible for SNAP does not necessarily mean access to sufficient, healthy food as the funds allocated to participants are not enough to provide food for the entire month they are meant for. Given a tight budget and mouths to feed, participants face the difficult choice of purchasing expensive fruits and vegetables that their family may not enjoy, or cheap, easy-to-prepare food like instant noodles and canned soups. 

This contributes to negative health outcomes such as obesity and diabetes, both of which are prevalent issues that should be addressed. These health issues are particularly concerning for marginalized populations, as they are more likely to experience other medical conditions such as heart disease due to additional stressors, physical, or environmental factors.[^1] However, seeing medical professionals can be incredibly costly, with visits to primary care doctors, medical specialists, and dentists racking up hundreds if not thousands of dollars without healthcare coverage. Families and individuals who are low-income or live below the poverty line often can not afford to buy private health insurance, relying on the possibility of public healthcare coverage from the government. 

[^1]: Gee, G., & Ford, C. (2011). STRUCTURAL RACISM AND HEALTH INEQUITIES: Old Issues, New Directions. Du Bois Review: Social Science Research on Race, 8(1), 115-132. <doi:10.1017/S1742058X11000130>

Unfortunately, policies for government programs like public health insurance and SNAP can vary heavily from state to state, in some part due to their political leaning. States are typically categorized as red (Republican Party) or blue (Democratic Party), and blue states are often portrayed in media as spending more money on public policies, including healthcare and SNAP costs. 

By researching the relationship between SNAP participation and public healthcare coverage we hope to discover if the people who need healthcare the most are able to receive it. Examining how SNAP participation has changed over time also provides valuable information on how effective the program is in providing support during challenging times in recent years like the pandemic. Additionally, we hope to find out more about what the demographics of these populations are and whether political standing plays a role in access to public assistance programs. To do this, we asked the following questions: Do states with the highest SNAP participants also reflect the highest public healthcare coverage? Does being a primarily red or blue state affect SNAP costs and public healthcare coverage? Which racial group participates in the SNAP program the most among different states? And when was the highest participation of SNAP within the last 5 years and is there an upward or downward trend?

We hope that through our research we can reach policymakers, government representatives, as well as healthcare professionals -- individuals whose decisions and actions shape access to public welfare. By looking into food assistance as well as its ties to healthcare coverage, this information will help figures in policy take more than just cost and a head count into consideration when budgeting and designing public programs. It is also important to note that it may seem like we are researching a small group of people within a much larger population. However, economic security is fragile and impacted by broader structural events, meaning that at any time our neighbors, friends, and even you can be hit by an unexpected rough patch and come to need support from public programs. Therefore it is beneficial to everyone to have strong, well-informed welfare systems in place. 

# **Data - Ella**

Our research is centered on uncovering the relationship between SNAP participation and healthcare coverage across different U.S. states. We specifically chose these datasets because they provide comprehensive state-level data, effectively representing the correlation we aim to analyze.

1.  Focus on State-Level Data: Both SNAP participation and healthcare coverage data were selected at the state level to analyze the potential correlation across the state level.

2.  Use of Latest Five-Year Data: To mitigate potential biases associated with data from any specific year, we decided to analyze data from the most recent five-year period. This approach allows for a more balanced and current understanding of the trends and patterns in SNAP participation and healthcare coverage.

3.  Selection of Public Healthcare Data: Considering that SNAP is a public assistance program, we specifically focused on public healthcare coverage data. This decision was made to align the healthcare data with the public nature of SNAP, enabling a more accurate comparison between these two aspects of public welfare.

4.  Inclusion of Demographic and Political Data: Enhancing our analysis, we included the latest demographic data related to SNAP participation. Additionally, we incorporated data on states' political affiliations (red/blue states) to investigate how SNAP participation and healthcare coverage may relate to a state's political stance. This multifaceted approach enables us to explore the interplay between SNAP participation, healthcare coverage, demographic factors, and political influences within states.

The 'SNAP7Years' dataset is sourced from the USDA Food & Nutrition Service. As a product of a U.S. government agency, it is publicly accessible and includes a comprehensive monthly statistical sample from state agencies. The dataset encompasses households that are either participating in or have been excluded from SNAP.   It includes a monthly-reviewed statistical sample of these households, carried out by state agencies to assess program participation and efficacy.

In addition, the 'Healthcare5Years' and 'SNAP2022Race' datasets are both compiled by the United States Census Bureau, specifically from the ACS 1-Year Estimates PUMS. These datasets are publicly available, adhering to the U.S. Census Bureau's policy of open data access. They provide a broad demographic overview, capturing data from both urban and rural populations across the U.S. and Puerto Rico. The samples consist of a subset of ACS respondents who offer detailed, anonymized responses. This rich data source is instrumental in creating custom tables and facilitating in-depth statistical analyses.

Lastly, the classification of states into 'Red' and 'Blue' categories is based on information from CNN, providing a political context to our analysis. This classification aids in understanding how political leanings might correlate with differences in SNAP participation and healthcare coverage across states.

Here are the sources of the data

SNAP7Years: [USDA Food & Nutrition Service](https://www.fns.usda.gov/pd/supplemental-nutrition-assistance-program-snap)

Healthcare5Years: [United States Census Bureau](https://data.census.gov/mdat/#/search?ds=ACSPUMS1Y2022&vv=PERNP&cv=HICOV&rv=ucgid&wt=PWGTP&g=0400000US01,02,04,05,06,08,09,10,11,12,13,15,16,17,18,19,20,21,22,23,24,25,26,27,28,29,30,31,32,33,34,35,36,37,38,39,40,41,42,44,45,46,47,48,49,50,51,53,54,55,56)

SNAP2022Race: [United States Census Bureau](https://data.census.gov/mdat/#/search?ds=ACSPUMS1Y2022&cv=RAC1P,FS&rv=ucgid&wt=WGTP&g=0400000US01,02,04,05,06,08,09,10,11,12,13,15,16,17,18,19,20,21,22,23,24,25,26,27,28,29,30,31,32,33,34,35,36,37,38,39,40,41,42,44,45,46,47,48,49,50,51,53,54,55,56)

Red/Blue States: [CNN](https://www.cnn.com/election/2020/results/president)

\
Our data sources are credible and dependable, yet we encountered some potential data quality issues during the process of merging and manipulating the data.

1.  Non-continuous data

    Initially, our aim was to analyze the most recent five years of data. However, we encountered a limitation as the Census healthcare dataset began in 2023 and did not include data for 2020. Conversely, the SNAP dataset commenced in 2022. Consequently, we merged the datasets for the years they had in common: 2022, 2021, 2019, 2018, and 2017. This presents a challenge because the years are not sequentially consistent, potentially affecting the continuity of our analysis. Nevertheless, our objective remained to examine the latest available data over a five-year span.

2.  Race variables

    The racial variables were not distinctly delineated, a reflection of the contemporary understanding that identity can be fluid across a spectrum. Therefore, we concentrated our analysis on the primary racial categories available in the dataset: White, Asian, African-American, and Indigenous peoples.\

After cleaning and merging the datasets, here are the details of our final dataset.

Each row in the final dataset represents state-specific data on public healthcare coverage, SNAP participation, and race for each year. This includes total population figures, those covered by public healthcare, SNAP household and individual participation, and racial breakdown of SNAP participants.

The merged dataset contains 306 rows and 21 columns. Key variables include State, Year, Total(Healthcare), public health insurance coverage details, SNAP participation figures, total cost, and racial demographics of SNAP participants.

The merged dataset exclusively for 2022's SNAP racial data resulted in NA values for race in other years. A total of 255 missing values were noted for SNAP2022Race variables, which aligned with our focus on the most recent year, ensuring these gaps did not affect the overall analysis. Apart from SNAP2022Race variables, all columns contained valid data.

The following are details for each variable in the final dataset:

-   State: The name of the state.

-   Year: The year for which the data is relevant.

-   Total(Healthcare): Total number of individuals in the state.

-   With health insurance coverage: Number of individuals with public health insurance coverage.

-   No health insurance coverage: Number of individuals without public health insurance coverage.

-   SNAP Household Participation: The number of households participating in SNAP.

-   SNAP Persons Participation: The number of persons participating in SNAP.

-   SNAP Total Cost: Total cost of SNAP in the state.

-   SNAP Cost Per Household: Average cost of SNAP per participating household.

-   SNAP Cost Per Persons: Average cost of SNAP per participating person.

-   Total(SNAP-Racebased): Total number of SNAP participants based on race.

-   White_alone: Number of SNAP participants who are White alone.

-   Black or African American alone: Number of SNAP participants who are Black or African American alone.

-   American Indian alone: Number of SNAP participants who are American Indian alone.

-   Alaska Native alone: Number of SNAP participants who are Alaska Native alone.

-   American Indian and Alaska Native tribes specified; or American Indian or Alaska Native, not specified and no other races: Number of SNAP participants in specified American Indian and Alaska Native tribes, or unspecified American Indian or Alaska Native tribes without other races.

-   Asian alone: Number of SNAP participants who are Asian alone.

-   Native Hawaiian and Other Pacific Islander alone: Number of SNAP participants who are Native Hawaiian and Other Pacific Islander alone.

-   Some other race alone: Number of SNAP participants who are of some other race alone.

-   Two or More Races: Number of SNAP participants who belong to two or more races.

# **Methods - Julia**

*Bolded italics*[^2] are newly created variables.\
**Bolded**[^3] are original variables from our data.

[^2]: Bolded italics are newly created variables.

[^3]: Bolded are original variables from our data.

To answer our first question "Do states with the highest SNAP participants also reflect the highest public healthcare coverage,"  we used the following variables: **State, Year, Total, With public health coverage, and SNAP Persons Participation**.

We took the average of the ***Total*** variable for each state to get the **Average Population** and added it to a new data frame named ***SNAP_healthcare***. We calculated the average of **SNAP Persons Participation** for each state, divided that by the **Average Population**, and multiplied it by 100 to get the average percentage of SNAP participation for each state. To compute what percentage of each state has healthcare coverage (on average), we divided **With public health coverage** by **Total** and multiplying it by 100, then assigned this computation to the variable ***average_healthcare_coverage***. Merging  ***SNAP_healthcare*** and ***average_healthcare_coverage*** created a new dataframe which we named ***merged_SNAP_healthcare*****.**

From the new data frame, ***merged_SNAP_healthcare*** we selected the following variables: **State, *Average_Healthcare_Coverage_Percentage*,** and ***snap_participation_percent***, renamed the variables to more concise labels, and put them into a new variable ***percentage*** to allow easier plotting.We reshaped ***percentage*** into long-form, with Healthcare Coverage and SNAP Participation corresponding to ***type***, and their values corresponding to ***percentage***. We then utilized the new long-form data frame to create a boxplot that shows the average healthcare coverage and SNAP participation across all the states, to visualize if there is a relationship between the two variables. 

To create the barplot that addresses whether states with the highest SNAP Participation also have the highest healthcare coverage. First, we ranked the states by SNAP Participation, then filtered out the top 10 states and worst 10 states to create two separate bar plots that would help us compare both spectrums and visualize if there was an association. We repeated the methodology with healthcare coverage even though it does not directly answer our first research question, we wanted to see if there was an association by comparing the highest and lowest healthcare coverage to SNAP participation.

The methodology for our second research question, "Does being primarily a red or blue state affect SNAP costs and public healthcare insurance," we selected the following variables: **Total**, **With public health coverage**, and **SNAP Cost Per Persons**. 

To compute the healthcare coverage ratio state by state, we created a new variable ***healthcarecover_ratio*** and assigned **With public health coverage** divided by **Total**. Additionally, we created ***Mean_Healthcare_Cover_Ratio*** by averaging ***healthcarecover_ratio***, ***Mean_SNAP_Cost_Per_Persons*** by averaging **SNAP Cost Per Persons**. To create the new variable **state_politics** we created and assigned two new vectors with states corresponding to a red or blue political party. We created these new variables to be able to compare the average between SNAP costs and public healthcare coverage. 

We reshaped the data frame into long-form data frame to create our boxplots. Since the main question focuses on examining the correlation between a state's political stance and its healthcare coverage and SNAP cost per person, we decided to eliminate outliers, specifically Alaska and Hawaii, where SNAP per person exceeded \$200.

For our third question, "Which racial group participates in the SNAP program the most amongst different states," we selected the following variables: **State, White_alone, Black or African American alone, American Indian alone, Alaska Native alone, Asian alone,** and **Native Hawaiian and Other Pacific Islander alone.** I decided to choose these race variables because they are socially known as the primary race groups as well as have more impactful statistics. Additionally, the data in these variables is only counting people who are SNAP participants. We then omitted rows that had missing values to only isolate the data we wanted.

We then reshaped the data into long-form and renamed the races to more concise names. We grouped the rows by state and ranked them within their state to identify which race has the highest number of SNAP participants. We then plotted the race with the highest participation onto a map. We also decided to compare the races and their SNAP Participation nationally by creating a bar plot.

(Written by Sabrin)\
Methods for the last research question: "When was the highest participation of SNAP within the last 5 years and is there an upward or downward trend?" I have used the merged data to filter the top 10 states that have the most participated snap in the last 5 years. To make measurements for upward and downward. I have not included any variables. I have created a variable top_10_states to merge the merged data to and filter the states I want.  I have used the processed data to select and summarize for the variable. I haven't done any technique to remove or add anything.

# **Results - Sabrin**

\*This code and plots are made by Angelic

```{r}
#I took the average of Total(population) for each state across the years to calculate the Average Population for each state. Then I added it to a new dataframe named SNAP_healthcare.
SNAP_healthcare <- merged_data %>%
  group_by(State) %>%
  summarise(`Average Population` = mean(`Total`))
```

```{r}
#I took the average of SNAP Persons Participation for each state across the years. Then divided that by the Average Population to get the percentage of average SNAP participation for each state.
SNAP_participation <- merged_data %>%
  group_by(State) %>%
  summarise(Average_SNAP_Persons = mean(`SNAP Persons Participation`))

SNAP_healthcare <- left_join(SNAP_healthcare, SNAP_participation, by = "State")

SNAP_healthcare$snap_participation_percent <- (SNAP_healthcare$Average_SNAP_Persons / SNAP_healthcare$`Average Population`) * 100

```

```{r}
#Calculating the percentage of healthcare coverage for each state, and then computing the average healthcare coverage across all the years.
healthcare_coverage <- merged_data %>%
  mutate(Healthcare_Coverage_Percentage = (`With public health coverage` / Total * 100))

average_healthcare_coverage <- healthcare_coverage %>%
  group_by(State) %>%
  summarise(Average_Healthcare_Coverage_Percentage = mean(Healthcare_Coverage_Percentage, na.rm = TRUE))
```

```{r}
#Merging the dataset with average SNAP participation percentage and the dataset with average healthcare coverage percentage.
merged_SNAP_healthcare <- merge(average_healthcare_coverage, SNAP_healthcare, by = "State")

percentage <- merged_SNAP_healthcare %>% 
  select(State, Average_Healthcare_Coverage_Percentage, snap_participation_percent) %>% 
  rename(`Healthcare Coverage` = Average_Healthcare_Coverage_Percentage,
         `SNAP Participation` = snap_participation_percent)
```

```{r}
#I reshaped the percentage data into long-form.
percentage_long <- percentage %>%
  pivot_longer(cols = -State, names_to = "type", values_to = "percentage")

```

```{r}
#Plotting a boxplot using long form
plotHealthSnap <- ggplot(percentage_long, aes(x = type, y = percentage, fill = type)) +
  geom_boxplot() +
  scale_fill_manual(values = c("Healthcare Coverage" = "skyblue", "SNAP Participation" = "lightcoral")) +
  labs(title = "Healthcare Coverage and SNAP Participation Percentage",
       x = "Type",
       y = "Percentage") +
  theme_light()
print(plotHealthSnap)  
```

```{r}
#Ranking the states by SNAP participation percentage.
merged_SNAP_healthcare <- merged_SNAP_healthcare %>%
  mutate(Rank_SNAP_Participation = rank(-snap_participation_percent))



top_10_states <- merged_SNAP_healthcare %>%
  filter(Rank_SNAP_Participation <= 10) %>%
  arrange(snap_participation_percent)

worst_10_states <- merged_SNAP_healthcare %>%
  filter(Rank_SNAP_Participation > n_distinct(State) - 10) %>%
  arrange(desc(snap_participation_percent))
```

In the Top 10 States bar plot, there appears to be a connection between SNAP Participation and Healthcare Coverage, with increased participation in SNAP aligning with higher healthcare coverage. Conversely, in the Worst 10 States, no apparent correlation is evident. In general, upon comparing the two bar graphs, it is clear that states with the highest SNAP participation do not consistently boast the highest healthcare coverage.

```{r}
#Graphing the top 10 and worst 10 states based on SNAP participation.
top_states <- percentage_long %>%
  filter(State %in% top_10_states$State)

ggplot(top_states, aes(x = State, y = percentage, fill = type)) +
  geom_bar(stat = "identity", position = "dodge", width = 0.7) +
  labs(title = "Top 10 States: SNAP Participation",
       x = "State", y = "Percentage") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_manual(values = c("Healthcare Coverage" = "skyblue", "SNAP Participation" = "lightcoral"))
```

```{r}
worst_states <- percentage_long %>%
  filter(State %in% worst_10_states$State)

ggplot(worst_states, aes(x = State, y = percentage, fill = type)) +
  geom_bar(stat = "identity", position = "dodge", width = 0.7) +
  labs(title = "Worst 10 States: SNAP Participation",
       x = "State", y = "Percentage") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_manual(values = c("Healthcare Coverage" = "skyblue", "SNAP Participation" = "lightcoral")) +
  scale_y_continuous(breaks = seq(0, 50, by = 10), limits = c(0, 50))
```

States with the greatest healthcare coverage may not always have the highest SNAP participation. Examining New Mexico, which has the highest coverage rates, reveals that its SNAP participation rates are nearly double those of Utah, which has the least coverage. It's intriguing that New Mexico tops both in SNAP participation and healthcare coverage, while Utah ranks lowest in both. One possible explanation is that Utah has a smaller population living below the poverty line, resulting in lower demand for SNAP and public healthcare coverage.

```{r}
#Ranking the states based on healthcare coverage percentage.
merged_SNAP_healthcare <- merged_SNAP_healthcare %>%
  mutate(Rank_Healthcare_Coverage = rank(-`Average_Healthcare_Coverage_Percentage`))

top_10_healthcare <- merged_SNAP_healthcare %>%
  filter(Rank_Healthcare_Coverage <= 10) %>%
  arrange(`Average_Healthcare_Coverage_Percentage`)

worst_10_healthcare <- merged_SNAP_healthcare %>%
  filter(Rank_Healthcare_Coverage > n_distinct(State) - 10) %>%
  arrange(desc(`Average_Healthcare_Coverage_Percentage`))
```

```{r}
#Graphing the top 10 and worst 10 states based on healthcare coverage.
top_healthcare_states <- percentage_long %>%
  filter(State %in% top_10_healthcare$State)

ggplot(top_healthcare_states, aes(x = State, y = percentage, fill = type)) +
  geom_bar(stat = "identity", position = "dodge", width = 0.7) +
  labs(title = "Top 10 States: Healthcare Coverage",
       x = "State", y = "Percentage") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_manual(values = c("Healthcare Coverage" = "skyblue", "SNAP Participation" = "lightcoral"))
```

```{r}
worst_healthcare_states <- percentage_long %>%
  filter(State %in% worst_10_healthcare$State)

ggplot(worst_healthcare_states, aes(x = State, y = percentage, fill = type)) +
  geom_bar(stat = "identity", position = "dodge", width = 0.7) +
  labs(title = "Worst 10 States: Healthcare Coverage",
       x = "State", y = "Percentage") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_manual(values = c("Healthcare Coverage" = "skyblue", "SNAP Participation" = "lightcoral")) +
  scale_y_continuous(breaks = seq(0, 50, by = 10), limits = c(0, 50))
```

\*This code and plots are made by Ella

```{r}
red_states <- c("Alabama", "Alaska", "Arkansas", "Florida", "Idaho", "Indiana", 
                "Iowa", "Kansas", "Kentucky", "Louisiana", "Mississippi", "Missouri", 
                "Montana", "Nebraska", "North Carolina", "North Dakota", "Ohio", 
                "Oklahoma", "South Carolina", "South Dakota", "Tennessee", "Texas", 
                "Utah", "West Virginia", "Wyoming")

blue_states <- c("Arizona", "California", "Colorado", "Connecticut", "Delaware", 
                 "District of Columbia", "Georgia", "Hawaii", "Illinois", "Maine", 
                 "Maryland", "Massachusetts", "Michigan", "Minnesota", "Nevada", 
                 "New Hampshire", "New Jersey", "New Mexico", "New York", "Oregon", 
                 "Pennsylvania", "Rhode Island", "Vermont", "Virginia", "Washington", 
                 "Wisconsin")


merged_data_Ella <- merged_data %>%
  select(State, Year, `Total`, `With public health coverage`,  `SNAP Cost Per Persons`, `SNAP Cost Per Household`) %>%
  mutate(healthcarecover_ratio = `With public health coverage` / `Total`,
         state_politic = case_when(
           State %in% red_states ~ "Red",
           State %in% blue_states ~ "Blue",
           TRUE ~ "Other" 
         )) %>%
  group_by(State, state_politic) %>%
  summarise(
    Mean_Healthcare_Cover_Ratio = mean(healthcarecover_ratio, na.rm = TRUE),
    Mean_SNAP_Cost_Per_Person = mean(`SNAP Cost Per Persons`, na.rm = TRUE),
    Mean_SNAP_Cost_Per_Household = mean(`SNAP Cost Per Household`),
    .groups = "drop" 
  )

```

Therefore, we would like to discover other factors that could potentially affect the SNAP cost and public healthcare coverage.

The box plot provided illustrates that, on average, Blue states exhibit a higher median for both SNAP cost per person and public healthcare coverage. While the median difference between Blue and Red states may appear minimal, it's noteworthy that Red states display a broader range for both variables. This suggests a greater disparity within Red states in terms of SNAP costs and public healthcare coverage.

The findings suggest that the political orientation of certain states could influence their policies regarding SNAP costs and public healthcare coverage. Such a trend is concerning, as these policies are closely tied to the general health and well-being of the population. It is recommended that policymakers consider healthcare coverage and programs like SNAP as essential, yet distinct from the standard policy agenda. This approach highlights the importance of these programs in the context of overall living conditions and underscores the need for them to be addressed with due consideration and independence from political biases.

```{r}


# Filter the data to include only SNAP cost per person under $200
filtered_data <- merged_data_Ella %>%
  filter(Mean_SNAP_Cost_Per_Person < 200)

# Reshape the data for plotting
long_data <- filtered_data %>%
  pivot_longer(cols = c("Mean_Healthcare_Cover_Ratio", "Mean_SNAP_Cost_Per_Person"),
               names_to = "Metric", values_to = "Value")

# Create the boxplot
ggplot(long_data, aes(x = state_politic, y = Value, fill = state_politic)) +
  geom_boxplot() +
  facet_wrap(~Metric, scales = "free_y", labeller = as_labeller(c(Mean_Healthcare_Cover_Ratio = "Healthcare Coverage (%)", Mean_SNAP_Cost_Per_Person = "SNAP Cost per Person ($)"))) +
  labs(title = "Healthcare Coverage and SNAP Cost Per Person by State Politics", x= "", y="") +
  scale_fill_manual(values = c("Red" = "red", "Blue" = "blue")) +
  theme_minimal()

```

\*This code and plots are made by Julia

Another factor we would like to analyze was demographics, especially Race data. 

This map displays the primary racial group with the highest SNAP participation in each state, excluding Alaska and Hawaii. In the Southern U.S. region, encompassing Louisiana, Mississippi, Alabama, Georgia, and South Carolina, the race with the highest participation is Black or African American. Notably, two states in New England, namely Maryland and Delaware, also share this same racial group with the highest participation

```{r}
##Calculations
#To find the indices and names of columns in the data set
printNameIndices <- function(dataSet) {
  colIndices <- 1:ncol(dataSet)
  colNames <- names(dataSet)
  colNamesIndices <- paste(colIndices, colNames, sep = ": ")
  cat(colNamesIndices, sep = "\n")
}

#To select the columns I need
selectedMergeRace <- merged_data %>%
  select(1, 13:16, 18, 19) %>%
  na.omit(merged_data)

#Reshape data into long form
reshapeSNAP <- selectedMergeRace %>%
  pivot_longer(
    cols = c(-State),
    names_to = "Race",
    values_to = "Snap_Count"
  )

#Rename values
reshapeSNAP <- reshapeSNAP %>%
  mutate(Race = recode(Race,
      "White_alone" = "White",
      "Black or African American alone" = "Black or African American",
      "Alaska Native alone" = "Alaska Native",
      "American Indian alone" = "Indigenous People",
      "Asian alone" = "Asian",
      "Native Hawaiian and Other Pacific Islander alone" = "Native Hawaiian or PI"))

#Ranking races by who uses the SNAP program the most within each state
rankRaceWithinStates <- reshapeSNAP %>%
  group_by(State) %>%
  arrange(desc(Snap_Count), .by_group = TRUE) %>%
  mutate(rank = row_number()) %>%
  arrange(rank)

#Grouping Snap_Count by Race
sumSNAPbyRace <- reshapeSNAP %>%
  group_by(Race) %>%
  mutate(SumSNAPcount = sum(Snap_Count)) %>%
  distinct(SumSNAPcount) %>%
  arrange(SumSNAPcount)
```

```{r}
##Map Plotting

library(maps)

states_map <- map_data("state")
states_map$region <- tolower(states_map$region)

#Consistency
rankRaceWithinStates$State <- tolower(rankRaceWithinStates$State)

#Create new data frame of the highest race in each state for mapping
highestRaceSNAP <- left_join(states_map, rankRaceWithinStates[1:51, ], by = c("region" = "State"))

#Plotting a US map with the race that has the highest SNAP participation
raceMap <- ggplot(highestRaceSNAP, aes(x = long, y = lat, group = group, fill = factor(Race))) +
  geom_polygon(color = "black") +
  theme_void() +
  labs(title = "Map of Races that have the Highest SNAP Participants",
       fill = "Race")
print(raceMap)
```

Additionally, when examining SNAP participation across races throughout the US, the White population has the highest participation rate, followed by Black or African American individuals.

```{r}
#Bar Plotting
#Plotting a Bar plot to compare the race groups nationally
raceBar <- ggplot(sumSNAPbyRace) +
  geom_col(aes(x = Race, y = SumSNAPcount, fill = Race)) +
  scale_fill_manual(values = c("red", "blue", "green", "orange", "purple","lightblue")) +
  labs(title = "SNAP Enrollment Across Racial Groups",
       x = "Race",
       y = "Number of Snap Participants") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10))
print(raceBar)
```

## 

\*This code and plots are made by Sabrin

The bar chart illustrates the upward and downward trends over the past five years for states with the highest SNAP participation. Each bar is labeled with the corresponding year, and its colors indicate whether the trend is upward or downward. This pertains specifically to the top 10 states with the highest SNAP participants.

```{r}
# I filtered the data for the highest 20 states. I just wanted to focus on 20 states.
top_10_states <- merged_data %>%
  group_by(State) %>%
  summarize(avg_participation = mean(`SNAP Household Participation`, na.rm = TRUE)) %>%
  top_n(10, wt = avg_participation) %>%
  pull(State)
```

```{r}
filtered_data <- merged_data %>%
  filter(State %in% top_10_states)
# Made in order levels year based on the SNAP participation average.
filtered_data$Year <- factor(filtered_data$Year, levels = unique(filtered_data$Year[order(-filtered_data$`SNAP Household Participation`)]))

```

```{r}
# Calculated the SNAP participation average for each state. 
avg_participation_per_state <- filtered_data %>%
  group_by(State) %>%
  summarize(avg_participation = mean(`SNAP Household Participation`, na.rm = TRUE))

```

```{r}
# I reordered the states based on the SNAP participation average.
filtered_data$State <- factor(filtered_data$State, levels = avg_participation_per_state$State[order(avg_participation_per_state$avg_participation)])
# Also I made a color palette with the same order as the levels of Year.
color_palette <- RColorBrewer::brewer.pal(5, "Set1")[order(levels(filtered_data$Year))]
```

```{r}
# Bar for SNAP participation in the highest 10 states
snap_bar_top_10 <- ggplot(filtered_data, aes(x = State, y = `SNAP Household Participation`, fill = Year)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "SNAP Participation in Top 10 States",
       x = "State",
       y = "SNAP Household Participation",
       fill = "Year") +
  scale_y_continuous(labels = scales::label_number_si()) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
  

print(snap_bar_top_10)

#The year label on the bar side represent the in order of the years how snap went upward and downward for the past five years with it's color. 

```

The connection between healthcare coverage and participation in the Supplemental Nutrition Assistance Program (SNAP) is intricate, demonstrating an irregular correlation. States with elevated SNAP involvement typically display superior healthcare coverage. For instance, Utah registers the lowest participation and coverage, while New Mexico leads in both. Analyzing contributing factors, we observed that blue states allocate more funds to SNAP and boast higher healthcare coverage than their red counterparts. Political inclinations, possibly linked to broader socioeconomic elements, contribute to variations among states. Although the Southern U.S. has a historical prevalence of Black or African American residents, this doesn't entirely elucidate the disparities. The findings underscore the necessity for targeted nationwide research and policies to ensure impartial access to vital services. Addressing systemic hurdles and recognizing distinct historical contexts is essential for fostering fairness across states and demographic categories.

# **Findings - Ella, Angelic, Julia**

The relationship between healthcare coverage and SNAP (Supplemental Nutrition Assistance Program) participation is complex and indicative of broader policy and socioeconomic dynamics. Across all the states plotted, the percentage of healthcare coverage was more than double of the percentage of SNAP participation. It is interesting that there is such a discrepancy between healthcare coverage and SNAP participation, it likely has something to do with the fact that medical expenses can be much more costly than food without health insurance and perhaps social stigma surrounding the use of "food stamps". Despite this consistent discrepancy, the findings do not reveal a clear correlation between the two variables . We observed that states with the highest SNAP participation tend to exhibit overall higher levels of comprehensive healthcare compared to the states with the lowest participation. For example, Utah shows the lowest for both SNAP participation and healthcare coverage while New Mexico has the highest SNAP participation and highest healthcare coverage. However, some of the states with the lowest SNAP participation have the same percentage of public healthcare coverage as states with the highest SNAP participation. 

To delve deeper into the relationship between these two variables, we considered various factors that might explain the stark disparity between certain states. Political leanings, for example, could indicate broader socioeconomic trends and public policy orientations, as blue states often show more investment in welfare programs. For example, New Mexico is a blue state and Utah is a red state, political preferences might play a role in explaining the differences among states.  Additionally, the demographic composition of the Southern U.S., with its historically higher Black or African American population, may explain the race most represented in SNAP participation. As seen on the map, a majority of the SNAP users are White, yet marginalized groups are more likely to experience obesity. Low SNAP participation may contribute to this however, we are still unsure if this is a direct association with access to healthcare access.

Initially, our goal was to explore a potential correlation between SNAP and healthcare coverage at the state level. While we didn't find a strong overall relationship, some states displayed notable patterns. These patterns might be partially explained by political affiliations or racial demographics, but it's challenging to assert a strong correlation between these factors and SNAP or healthcare coverage.

However, this analysis has limitations. Categorizing states strictly by political stance can lead to oversimplification and potentially overlook nuances, especially in swing states. Also, the statistical data from states might not fully reflect the current health research and outcomes for marginalized groups.

For future studies, it is advisable to include more complex socioeconomic variables and medical data to discern clearer patterns in the relationship between SNAP and healthcare coverage. Continued focus on identifying those in need of SNAP participation and public insurance coverage is crucial. By doing so, we can enhance the overall quality of public policy, ensuring more individuals benefit from these programs.

# **Summary - Angelic**

The report investigates the intricate relationship between SNAP participation and healthcare coverage across U.S. states, aiming to understand demographic trends, changes in demand for food assistance, and potential political and demographic influences. Utilizing state-level data from sources like the USDA and U.S. Census Bureau, the analysis reveals disparities in access to welfare programs. 

While no clear correlation emerged between SNAP participation and healthcare coverage across states, our findings emphasized that states with higher SNAP participation often showcased better healthcare coverage. Factors like political affiliations, racial demographics, and socioeconomic variations were explored but couldn't entirely explain the observed disparities. This calls for policy interventions to ensure equitable access to essential services, independent of political biases, to improve public well-being comprehensively. We hope that in the future our research can be built upon by looking at how factors such as poverty levels, participation in private healthcare coverage, ease of application to public programs, and comparing the eligibility requirements between public healthcare and SNAP, to create a more nuanced understanding of our findings. 

Our holistic approach to well-being extends beyond looking at participation in just one public assistance program and exploring more than just general variables like cost to understand the SNAP program. By seeking demographic information and social influences, we are advocating for figures in policy and government programs to start thinking more about who their program serves and take on a more collective framework to ensure everyone has access to nutritious food and medical services.
