---
title: "Final Project"
output: html_document
date: "2023-11-30"
---


# Data Documentation

## 1. Link for each datasets

-   Healthcare5Years: This is the link for 2022 setting, got each year data then merged (2022 2021 2019 2018 2017)\
    <https://data.census.gov/mdat/#/search?ds=ACSPUMS1Y2022&vv=PERNP&cv=HICOV&rv=ucgid&wt=PWGTP&g=0400000US01,02,04,05,06,08,09,10,11,12,13,15,16,17,18,19,20,21,22,23,24,25,26,27,28,29,30,31,32,33,34,35,36,37,38,39,40,41,42,44,45,46,47,48,49,50,51,53,54,55,56>

-   SNAP2022Race\
    <https://data.census.gov/mdat/#/search?ds=ACSPUMS1Y2022&cv=RAC1P,FS&rv=ucgid&wt=WGTP&g=0400000US01,02,04,05,06,08,09,10,11,12,13,15,16,17,18,19,20,21,22,23,24,25,26,27,28,29,30,31,32,33,34,35,36,37,38,39,40,41,42,44,45,46,47,48,49,50,51,53,54,55,56>

-   SNAP7Years\
    <https://www.fns.usda.gov/pd/supplemental-nutrition-assistance-program-snap>

-   Red/Blue states\
    <https://en.wikipedia.org/wiki/2020_United_States_presidential_election>

## 2. Explanation of all variables

1.  SNAP2022Race: SNAP Participant race categorization in 2022
    a.  SNAP : Yes/No \*Only included Yes since this analysis is focused on SNAP participats
    b.  Total : Total participates count
    c.  White alone
    d.  Black or African American alone
    e.  American Indian alone
    f.  Alaska Native alone
    g.  American Indian and Alaska Native tribes specified; or American Indian or Alaska Native, not specified and no other races
    h.  Asian alone
    i.  Native Hawaiian and Other Pacific Islander alone
    j.  Some other race alone
    k.  Two or More Races
2.  Healthcare5Years : Total healthcare coverage in 2022 2021 2019 2018 2017
    a.  Total(Healthcare): Total population
    b.  With health insurance coverage: Population with insurance
    c.  No health insurance coverage: Population without insurance
3.  SNAP7Years : SNAP participants in 2023 2022 2021 2020 2019 2018 2017
    a.  SNAP Household Participation : Total household count participating SNAP
    b.  SNAP Persons Participation : Total person count participating SNAP
    c.  SNAP Total Cost : Total amount spend in SNAP
    d.  SNAP Cost Per Household
    e.  SNAP Cost Per Persons

## 3. Basic summary statistics

This merged dataset comprises 306 rows and 21 columns. It contains a total of 255 missing values across all SNAP2022Race variables, due to the exclusive merging of 2022 SNAP Race data. However, as our objective with the Race variable is to examine the most recent year, which is 2022, these missing values do not compromise the analysis. Except for the SNAP2022Race variables, all the columns have a valid data.\

```{r echo=FALSE}
library(tidyverse)

SNAP2022Race <- read_delim("../Datasets/SNAP_Race_2022.csv")
SNAP2022Race <- rename(SNAP2022Race, State = `Selected Geographies`)
head(SNAP2022Race,5)

Healthcare5Years <- read_delim("../Datasets/Healthcare_Total.csv")
Healthcare5Years <- rename(Healthcare5Years, State = `Selected Geographies`)
head(Healthcare5Years,5)

SNAP7Years <- read_delim("../Datasets/SNAP_7years.csv")
head(SNAP7Years,5)


merged_data <- inner_join(Healthcare5Years, SNAP7Years, by = c("State", "Year"))
merged_data <- inner_join(merged_data, SNAP2022Race, by = c("State", "Year"))


head(merged_data,5)

dim(merged_data)

missing_or_invalid_values_count <- merged_data %>%
  summarise_all(~sum(is.na(.) | . == ""))

missing_or_invalid_values_count
 
```

3.  

## Ella

**Main Question**\
Does being a primarily red or blue state affect SNAP Cost and healthcare coverage?

It's widely believed that blue states allocate more funds to public policies like healthcare and SNAP. This analysis aims to verify whether this belief holds for healthcare coverage and SNAP expenses, with the intention of providing policymakers and other stakeholders with data-driven recommendations based on our findings.

\
In the context of the most recent presidential election in the United States, states are often classified as "Red" or "Blue" based on the predominant voting preference for the Republican or Democratic party, respectively.

\<Red States (Republican - Donald Trump)\>

Alabama,Alaska,Arkansas,Florida,Idaho,Indiana,Iowa,Kansas,Kentucky,Louisiana,Mississippi,Missouri,Montana,Nebraska,North Carolina,North Dakota,Ohio,Oklahoma,South Carolina,South Dakota,Tennessee,Texas,Utah,West Virginia,Wyoming

\<Blue States (Democratic - Joe Biden)\>

Arizona,California,Colorado,Connecticut,Delaware,District of Columbia,Georgia,Hawaii,Illinois,Maine,Maryland ,Massachusetts,Michigan,Minnesota,Nevada,New Hampshire,New Jersey,New Mexico,New York,Oregon,Pennsylvania,Rhode Island,Vermont,Virginia,Washington,Wisconsin

Additionally, my analysis is centered on state-level comparisons. To facilitate this, I computed the average values for key variables: health insurance coverage, and SNAP Cost Per Person. Considering the varying population sizes across states, I also calculated the ratio of healthcare insurance coverage to provide a more balanced comparison.

```{r echo=FALSE}

red_states <- c("Alabama", "Alaska", "Arkansas", "Florida", "Idaho", "Indiana", 
                "Iowa", "Kansas", "Kentucky", "Louisiana", "Mississippi", "Missouri", 
                "Montana", "Nebraska", "North Carolina", "North Dakota", "Ohio", 
                "Oklahoma", "South Carolina", "South Dakota", "Tennessee", "Texas", 
                "Utah", "West Virginia", "Wyoming")

blue_states <- c("Arizona", "California", "Colorado", "Connecticut", "Delaware", 
                 "District of Columbia", "Georgia", "Hawaii", "Illinois", "Maine", 
                 "Maryland", "Massachusetts", "Michigan", "Minnesota", "Nevada", 
                 "New Hampshire", "New Jersey", "New Mexico", "New York", "Oregon", 
                 "Pennsylvania", "Rhode Island", "Vermont", "Virginia", "Washington", 
                 "Wisconsin")


merged_data_Ella <- merged_data %>%
  select(State, Year, `Total(Healthcare)`, `With health insurance coverage`,  `SNAP Cost Per Persons`, `SNAP Cost Per Household`) %>%
  mutate(healthcarecover_ratio = `With health insurance coverage` / `Total(Healthcare)`,
         state_politic = case_when(
           State %in% red_states ~ "Red",
           State %in% blue_states ~ "Blue",
           TRUE ~ "Other" 
         )) %>%
  group_by(State, state_politic) %>%
  summarise(
    Mean_Healthcare_Cover_Ratio = mean(healthcarecover_ratio, na.rm = TRUE),
    Mean_SNAP_Cost_Per_Person = mean(`SNAP Cost Per Persons`, na.rm = TRUE),
    Mean_SNAP_Cost_Per_Household = mean(`SNAP Cost Per Household`),
    .groups = "drop" 
  )

head(merged_data_Ella,5)

```

Firstly, an analysis is conducted to compare the average healthcare coverage ratio with the average SNAP cost per person, grouped by whether states are predominantly Red or Blue.

Upon examining the initial plot, it becomes evident that certain outlier values obscure the underlying trends within the data. To obtain a clearer view of the patterns, the dataset will be refined by excluding entries where the average SNAP cost per person exceeds 200.\

```{r echo=FALSE}

# Reshape the data for plotting
long_data <- merged_data_Ella %>%
  pivot_longer(cols = c("Mean_Healthcare_Cover_Ratio", "Mean_SNAP_Cost_Per_Person"),
               names_to = "Metric", values_to = "Value")

# Create the boxplot
ggplot(long_data, aes(x = state_politic, y = Value, fill = state_politic)) +
  geom_boxplot() +
  facet_wrap(~Metric, scales = "free_y") +
  labs(title = "Healthcare Coverage and SNAP Cost Per Person by State Politics", x= "", y="") +
  scale_fill_manual(values = c("Red" = "red", "Blue" = "blue")) +
  theme_minimal()


```

Once the outliers, specifically Alaska and Hawaii, were excluded, the data pattern became more discernible.

\
**Healthcare Coverage Ratio:** Blue states appear to have a higher median healthcare coverage ratio than Red states. The data for Blue states is more tightly clustered, indicating less variation in healthcare coverage ratios within these states. Red states show more variability, indicating some Red states with significantly lower healthcare coverage ratios.

**SNAP Cost Per Person:** The median SNAP cost per person is roughly the same for both Blue and Red states, but the range is broader for Red states, suggesting a greater disparity in SNAP costs within these states.

Although the average differences are not statistically significant, still there is a gap between the blue and red states, and even the distribution indicates some disparities. Blue states typically have higher healthcare coverage and SNAP expenditures per person than red states, suggesting that a state's political leanings may influence policies related to fundamental needs like food and healthcare. While our analysis did not employ extensive statistical methods and political affiliation is not the sole determining factor, these findings provide a useful perspective for state policymakers to consider.

```{r echo=FALSE}


# Filter the data to include only SNAP cost per person under $200
filtered_data <- merged_data_Ella %>%
  filter(Mean_SNAP_Cost_Per_Person < 200)

# Reshape the data for plotting
long_data <- filtered_data %>%
  pivot_longer(cols = c("Mean_Healthcare_Cover_Ratio", "Mean_SNAP_Cost_Per_Person"),
               names_to = "Metric", values_to = "Value")

# Create the boxplot
ggplot(long_data, aes(x = state_politic, y = Value, fill = state_politic)) +
  geom_boxplot() +
  facet_wrap(~Metric, scales = "free_y", labeller = as_labeller(c(Mean_Healthcare_Cover_Ratio = "Healthcare Coverage (%)", Mean_SNAP_Cost_Per_Person = "SNAP Cost per Person ($)"))) +
  labs(title = "Healthcare Coverage and SNAP Cost Per Person by State Politics", x= "", y="") +
  scale_fill_manual(values = c("Red" = "red", "Blue" = "blue")) +
  theme_minimal()

```

## Julia

**Research Question**\
Which racial group participates in the SNAP program the most amongst different states?

```{r echo=FALSE}
##Calculations
#To find the indices and names of columns in the data set
printNameIndices <- function(dataSet) {
  colIndices <- 1:ncol(dataSet)
  colNames <- names(dataSet)
  colNamesIndices <- paste(colIndices, colNames, sep = ": ")
  cat(colNamesIndices, sep = "\n")
}

#To select the columns I need
selectedMergeRace <- merged_data %>%
  select(1, 13:16, 18, 19) %>%
  na.omit(merged_data)

#Reshape data into long form
reshapeSNAP <- selectedMergeRace %>%
  pivot_longer(
    cols = c(-State),
    names_to = "Race",
    values_to = "Snap_Count"
  )

#Rename values
reshapeSNAP <- reshapeSNAP %>%
  mutate(Race = recode(Race,
      "White_alone" = "White",
      "Black or African American alone" = "Black or African American",
      "Alaska Native alone" = "Alaska Native",
      "American Indian alone" = "Indigenous People",
      "Asian alone" = "Asian",
      "Native Hawaiian and Other Pacific Islander alone" = "Native Hawaiian or PI"))

#Ranking races by who uses the SNAP program the most within each state
rankRaceWithinStates <- reshapeSNAP %>%
  group_by(State) %>%
  arrange(desc(Snap_Count), .by_group = TRUE) %>%
  mutate(rank = row_number()) %>%
  arrange(rank)

#Grouping Snap_Count by Race
sumSNAPbyRace <- reshapeSNAP %>%
  group_by(Race) %>%
  mutate(SumSNAPcount = sum(Snap_Count)) %>%
  distinct(SumSNAPcount) %>%
  arrange(SumSNAPcount)
```

```{r echo=FALSE}
##Map Plotting

library(maps)

states_map <- map_data("state")
states_map$region <- tolower(states_map$region)

#Consistency
rankRaceWithinStates$State <- tolower(rankRaceWithinStates$State)

#Create new data frame of the highest race in each state for mapping
highestRaceSNAP <- left_join(states_map, rankRaceWithinStates[1:51, ], by = c("region" = "State"))

#Plotting a US map with the race that has the highest SNAP participation
raceMap <- ggplot(highestRaceSNAP, aes(x = long, y = lat, group = group, fill = factor(Race))) +
  geom_polygon(color = "black") +
  theme_void() +
  labs(title = "Map of Races that have the Highest SNAP Participants",
       fill = "Race")
print(raceMap)
```

```{r}
#Bar Plotting
#Plotting a Bar plot to compare the race groups nationally
raceBar <- ggplot(sumSNAPbyRace) +
  geom_col(aes(x = Race, y = SumSNAPcount, fill = Race)) +
  scale_fill_manual(values = c("red", "blue", "green", "orange", "purple","lightblue")) +
  labs(title = "SNAP Enrollment Across Racial Groups",
       x = "Race",
       y = "Number of Snap Participants") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10))
print(raceBar)
```

## Angelic

**Research Question**

### Do states with the highest SNAP participation also reflect the highest healthcare coverage?

```{r echo=FALSE, include=FALSE}
head(merged_data)
```

I am calculating the average individual SNAP participation for each state across all the years (2017-2022, missing 2020). Then, I am ranking the states based on average SNAP participation.

```{r}
average_snap_participation <- merged_data %>%
  group_by(State) %>%
  summarise(Average_SNAP_Participation = mean(`SNAP Persons Participation`, na.rm = TRUE))

head(average_snap_participation)

ranked_states <- average_snap_participation %>%
  arrange(desc(Average_SNAP_Participation)) %>%
  mutate(SNAP_Participation_Rank = rank(-Average_SNAP_Participation, na.last = "keep"))

ranked_states
```

Calculating the percentage of healthcare coverage for each state, and then computing the average healthcare coverage across all the years.

```{r}
healthcare_coverage <- merged_data %>%
  mutate(Healthcare_Coverage_Percentage = (`With health insurance coverage` / `Total(Healthcare)` * 100))

average_healthcare_coverage <- healthcare_coverage %>%
  group_by(State) %>%
  summarise(Average_Healthcare_Coverage_Percentage = mean(Healthcare_Coverage_Percentage, na.rm = TRUE))

head(average_healthcare_coverage)
```

Merging the dataset with SNAP Participation rank and the dataset with average healthcare coverage percentage.

```{r}
merged_SNAP_healthcare <- merge(average_healthcare_coverage, ranked_states, by = "State")

head(merged_SNAP_healthcare)
```

I filtered out the top 10 ranked states (w.r.t. average SNAP participation) and created a bar graph that compare the rank to healthcare coverage percentage.

```{r}
top_10_states <- merged_SNAP_healthcare %>%
  filter(SNAP_Participation_Rank <= 10)

ggplot(top_10_states, aes(x = reorder(State, SNAP_Participation_Rank), y = Average_Healthcare_Coverage_Percentage)) +
  geom_bar(stat = "identity", fill = "skyblue",  width = 0.6) +
  geom_text(aes(label = paste0("#", SNAP_Participation_Rank)), vjust = -0.5, size = 3.5) +
  labs(x = "State", y = "Healthcare Coverage Percentage", title = "Top 10 Ranked States: SNAP Participation vs Healthcare Coverage") +
  coord_flip()
```

**The states were ranked by average SNAP participation across 5 years, with California having the highest participation at #1.**

I did the same thing but with the 10 worst ranked states (w.r.t. average SNAP participation).

```{r}
worst_10_states <- merged_SNAP_healthcare %>%
  filter(SNAP_Participation_Rank > (nrow(merged_SNAP_healthcare) - 10))

ggplot(worst_10_states, aes(x = reorder(State, SNAP_Participation_Rank), y = Average_Healthcare_Coverage_Percentage)) +
  geom_bar(stat = "identity", fill = "lightcoral", width = 0.6) +
  geom_text(aes(label = paste0("#", SNAP_Participation_Rank)), vjust = -0.5, size = 3.5) +
  labs(x = "State", y = "Healthcare Coverage Percentage", title = "10 Worst Ranked States: SNAP Participation vs Healthcare Coverage") +
  coord_flip()
```

**Surprisingly, the 10 states with the lowest average SNAP participation seem to actually have higher overall healthcare coverage when compared to the 10 states with the highest average SNAP participation**

First, I am taking the average of Total(Healthcare) for each state across the years to get an approximate value for the average total population of each state. Then I am creating a scatterplot that visualizes the relationship between SNAP participation and healthcare coverage, with the size of the points corresponding to the average total population of the states.

```{r}
average_population <- merged_data %>%
  group_by(State) %>%
  summarise(Average_Population = mean(`Total(Healthcare)`, na.rm = TRUE))

head(average_population)

merged_SNAP_healthcare <- merge(merged_SNAP_healthcare, average_population, by = "State")

head(merged_SNAP_healthcare)
```

```{r}
top_10_states <- merged_SNAP_healthcare %>%
  filter(SNAP_Participation_Rank <= 10)

ggplot(top_10_states, aes(x = Average_SNAP_Participation, y = Average_Healthcare_Coverage_Percentage, size = Average_Population, color = State)) +
  geom_point(alpha = 0.6) +
  scale_size(range = c(3, 15)) +
  labs(x = "Average SNAP Participation", y = "Average Healthcare Coverage Percentage", size = "Average Population", color = "State") +
  theme_minimal()
```

```{r}
worst_10_states <- merged_SNAP_healthcare %>%
  filter(SNAP_Participation_Rank > (nrow(merged_SNAP_healthcare) - 10))

ggplot(worst_10_states, aes(x = Average_SNAP_Participation, y = Average_Healthcare_Coverage_Percentage, size = Average_Population, color = State)) +
  geom_point(alpha = 0.6) +
  scale_size(range = c(3, 15)) +
  labs(x = "Average SNAP Participation", y = "Average Healthcare Coverage Percentage", size = "Average Population", color = "State") +
  theme_minimal()
```
## Sabrin
** research question
When was the highest participation of SNAP within the last  years and is there an upward or downward trend in the states?
```{r}
# I filtered the data for the highest 20 states. I just wanted to focus on 20 states.
top_10_states <- merged_data %>%
  group_by(State) %>%
  summarize(avg_participation = mean(`SNAP Household Participation`, na.rm = TRUE)) %>%
  top_n(10, wt = avg_participation) %>%
  pull(State)
```

```{r}
filtered_data <- merged_data %>%
  filter(State %in% top_10_states)
# Made in order levels year based on the SNAP participation average.
filtered_data$Year <- factor(filtered_data$Year, levels = unique(filtered_data$Year[order(-filtered_data$`SNAP Household Participation`)]))

```

```{r}
# Calculated the SNAP participation average for each state. 
avg_participation_per_state <- filtered_data %>%
  group_by(State) %>%
  summarize(avg_participation = mean(`SNAP Household Participation`, na.rm = TRUE))

```

```{r}
# I reordered the states based on the SNAP participation average.
filtered_data$State <- factor(filtered_data$State, levels = avg_participation_per_state$State[order(avg_participation_per_state$avg_participation)])
# Also I made a color palette with the same order as the levels of Year.
color_palette <- RColorBrewer::brewer.pal(5, "Set1")[order(levels(filtered_data$Year))]
```

```{r}
# Bar for SNAP participation in the highest 10 states
snap_bar_top_10 <- ggplot(filtered_data, aes(x = State, y = `SNAP Household Participation`, fill = Year)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "SNAP Participation in Top 10 States",
       x = "State",
       y = "SNAP Household Participation",
       fill = "Year") +
  scale_y_continuous(labels = scales::label_number_si()) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
  

print(snap_bar_top_10)

#The year label on the bar side represent the in order of the years how snap went upward and downward for the past five years with it's color. 

```
